Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-06 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.Rmd
add some references and change some language

title: "Clinical prediction with localized modeling using similarity-based cohorts: A systematic scoping review"
titlerunning: "Clinical prediction with localized modeling: A scoping review"

@@

A main driver of the development of clinical information systems (CIS) and \hl{computerized clinical decision support (CCDS)} tools has been to leverage the advantages of routinely collected health data toward improving care\hl{ [@]}.
These advantages include their immediate accessibility through the institutional \hl{electronic health record (EHR)}, billing database, or other source; their specificity to the institution and the population it serves; and the closeness in time of the available data.
Though by definition not collected for research use, the secondary research use of routinely collected health data has been put forth, as "practice-based evidence", a complement to the paradigm of evidence-based practice\hl{ [@]}.

Over the same period of time, interest in individualizing care from population-derived evidence-based recommendations to specific patient needs has driven the application of advanced computational tools\hl{ [@]}, including artificial intelligence (AI). Whereas classical models produced formulae involving a limited set of data elements that would be uniformly applied to all patients, AI models often process hundreds or more variables in opaque ways to yield predictions that depend on not only their values but their associations and interplay with each other\hl{ [@]}. The loss of straightforward interpretations of these models has limited their practical uptake and motivated the construction of explanatory statistics for opaque models as well as the development of more interpretable complex models\hl{ [@]}.

One of few methodologies to address all three of these established needs is one we call _localized modeling_, which \hl{we find in this review} to have been introduced several times independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that relies on a patient similarity measure[^psm] to extract a cohort of past or training cases that then inform the diagnosis, prognosis, or care of a new or test case.
Traditional CBR returns these retrieved cases to inform human decision-making\hl{ [@]}; the related nearest neighbors (NN) technique generates predictions from cohorts automatically via averaging (regression) or voting (classification) of their outcomes.
In contrast, localized modeling fits families of predictive models, for example generalized linear models (GLMs), to retrieved similarity cohorts in order to generate predictions.
\hl{While many ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions [@Molnar], these quantifications are not directly interpretable model components analogous to the split nodes of a DT or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected.}

@@

[^psm]: Similarity measures between more granular units of analysis, such as encounters and decision points, are often still referred to as patient similarity measures\hl{ [@]}.

@@

\hl{We also note that a large variety of AI tools, besides NN, use similarity measures or other relevance-based retrieval methods. Implicitly, back-propagation in ANNs relies on differentiation, which requires that all network nodes' activation functions and, therefore, the final model are continuous and differentiable, i.e. that small distances between cases yield small changes in prediction. More explicitly, kernel-based methods **TODO**. The key difference in localized modeling is that the retrieved cohort is treated as a component of the model and made available to the user. Indeed, retrieval procedures were the original algorithmic accomplishments of CBR [@] and the retrieved cohorts may be reused for numerous modeling tasks.}
Thus, localized models may embody both styles of CBR.

@@

Another interesting pattern lay in the years of publication (Figure \ref{fig:year}). While these studies trace back to the late 1990s, they have become more common\hl{ [@]}, which suggests that this is an active, though not rapidly expanding, approach. We note that CBR in health and medicine originated as early as 1990, and localized modeling emerged soon after CBR had established itself; the idea has been "in the air" for as long as CBR has been in use.

@@

Other studies addressed limitations of available tools. @Lopez2011 implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop more generalizable and data-agnostic clinical support tools [@Elter2007; @Liang2015; @Ng2015; @Zhang2018].
However, more informatical and legal challenges to implementation, including interoperability of systems and data and regulations concerning security and privacy, were rarely addressed.
We discuss this further in Section \ref{feasibility}.

@@

## Directions and expectations for future work

>>>>

title: "Clinical prediction using similarity cohort--based models: A systematic scoping review"
titlerunning: "Clinical prediction using similarity cohort--based models"

@@

A main driver of the development of clinical information systems (CIS) and \hl{computerized clinical decision support (CCDS)} tools has been to leverage the advantages of routinely collected health data toward improving care\hl{ }[@].
These advantages include their immediate accessibility through the institutional \hl{electronic health record (EHR)}, billing database, or other source; their specificity to the institution and the population it serves; and the closeness in time of the available data.
\hl{Though not} collected for research use, the secondary research use of routinely collected health data has been put forth, as "practice-based evidence", a complement to the paradigm of \hl{evidence-based practice }[@Holmqvist2015].

Over the same period of time, interest in individualizing care from population-derived evidence-based recommendations to specific patient needs has driven the application of advanced computational tools\hl{ }[@], including artificial intelligence (AI). Whereas classical models produced formulae involving a limited set of data elements that would be uniformly applied to all patients, AI models often process hundreds or more variables in opaque ways to yield predictions that depend on not only their values but their associations and interplay with each other\hl{ }[@]. The loss of straightforward interpretations of these models has limited their practical uptake and motivated the construction of explanatory statistics for opaque models as well as the development of more interpretable complex models\hl{ }[@].

One of few methodologies to address all three of these established needs is one we call _localized modeling_, which \hl{we find in this review} to have been introduced several times independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that relies on a \hl{patient similarity measure }[@Brown2016] to extract a cohort of past or training cases that then inform the diagnosis, prognosis, or care of a new or test case.[^psm]
\hl{Some of the earliest and simplest CBR tools returned} these retrieved cases \hl{ to inform human decision-making}[@Aamodt1994]; the related nearest neighbors (NN) technique generates predictions from cohorts automatically via averaging (regression) or voting (classification) of their outcomes.
In contrast, localized modeling fits families of predictive models, for example generalized linear models (GLMs), to retrieved similarity cohorts in order to generate predictions.
\hl{While many ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions }[@Molnar2023]\hl{, these quantifications are not directly interpretable model components analogous to the split nodes of a DT or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected.}

@@

[^psm]: Similarity measures between more granular units of analysis, such as encounters and decision points\hl{, or between patients' time series}, are often still referred to as patient similarity measures\hl{ in our sample}.

@@

\hl{We also note that a large variety of AI tools, besides NN, use similarity measures or other relevance-based retrieval methods. Implicitly, back-propagation in ANNs relies on differentiation, which requires that all network nodes' activation functions and, therefore, the final model are continuous and differentiable, i.e. that small distances between cases yield small changes in prediction. More explicitly, kernel-based methods **TODO**. The key difference in localized modeling is that the retrieved cohort is treated as a component of the model and made available to the user. Indeed, retrieval procedures were the most significant algorithmic requirements and accomplishments of CBR }[@Kolodner1992; @Gierl1998]\hl{ and the retrieved cohorts may be reused for numerous modeling tasks.}
Thus, localized models may embody both styles of CBR.

@@

Another interesting pattern lay in the years of publication (Figure \ref{fig:year}). While these studies trace back to the late 1990s, they have become \hl{slightly }more common, which suggests that this is an active, though not rapidly expanding, approach. We note that CBR in health and medicine originated as early as 1990, and localized modeling emerged soon after CBR had established itself; the idea has been "in the air" for as long as CBR has been in use.

@@

Other studies addressed limitations of available tools. @Lopez2011 implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop more generalizable and data-agnostic clinical support tools [@Elter2007; @Liang2015; @Ng2015; @Zhang2018].
However, more informatical and legal challenges to implementation, including interoperability of systems and data and regulations concerning security and privacy, were rarely addressed.
We discuss this further in Section \ref{future-work}.

@@

## Future work
